Proofs as Programs in Classical Logic\\Notes

### 2021-10-20 Notes/notes.tex
equality first thoughts

@@

\hspace*{-1cm}
\DisplayProof
}
\vspace{3cm}
\small{
	\AxiomC{}
	\RightLabel{\scriptsize Axiom}
	\UnaryInfC{$\vdash \alpha : \Pi_x\Sigma_y f(y)=x$}
	\RightLabel{\scriptsize $\forall E$}
	\UnaryInfC{$\vdash \alpha(x): \Sigma_y f(y)=x$}
	\AxiomC{}
	\RightLabel{\scriptsize Axiom}
	\UnaryInfC{$\vdash \beta : \Pi_x(P(x)\supset Q(f(x), x))$}
	\RightLabel{\scriptsize $\forall E$}
	\UnaryInfC{$\vdash \beta(y) : P(y)\supset Q(f(y), y)$}
	\AxiomC{}
	\RightLabel{\scriptsize Axiom}
	\UnaryInfC{$\vdash\Gamma\Pi_x P(x)$}
	\RightLabel{\scriptsize $\forall E$}
	\UnaryInfC{$\vdash\gamma(y) : P(y)$}
	\RightLabel{\scriptsize $\supset\hspace{-2pt}E$}
	\BinaryInfC{$\vdash \beta(y)(\gamma(y)) : Q(f(y), y)$}
	\AxiomC{}
	\RightLabel{\scriptsize u}
	\UnaryInfC{[$\vdash\delta : f(y) = x$]}
	\RightLabel{\scriptsize subst}
	\BinaryInfC{$\vdash \beta(y)(\gamma(y))_\delta : Q(x, y)$}
	\RightLabel{\scriptsize $\supset\hspace{-2pt}I^u$}
	\UnaryInfC{$\delta : f(y)=x\vdash\beta(y)(\gamma(y))_\delta : Q(x, y)$}
	\RightLabel{\scriptsize$\exists I$}
	\UnaryInfC{$\delta : f(y)=x\vdash \langle y, \beta(y)(\gamma(y))_\delta\rangle :\Sigma_y Q(x, y)$}
	\RightLabel{\scriptsize $\exists E$}
	\BinaryInfC{$\vdash \textbf{let }\langle y, \delta\rangle = \alpha(x)\textbf{ in }\langle y,\beta(y)(\gamma(y)\delta)\rangle: \Sigma_y Q(x, y)$}
	\RightLabel{\scriptsize $\forall I$}
	\UnaryInfC{$\lambda x.\textbf{let }\langle y, \delta\rangle = \alpha(x)\textbf{ in }\langle y,\beta(y)(\gamma(y))_\delta\rangle:\Pi_x \Sigma_y Q(x, y)$}

	\hspace*{-1cm}
	\DisplayProof
}
\hfill
\\\\\hspace{2cm}We can reduce the last term to $\lambda x.\langle \alpha(x)_1, \beta(\alpha(x)_1)(\gamma(\alpha(x)_1)_{\alpha(x)_2}\rangle$.

>>>>

\section{Type theory}

As we are interested in a proof-as-programs correspondence for first order logic our underlying type theory has to be of dependant nature, meaning that types can and will depend on terms.

One interesting aspect of choosing a type theory is how to handle equality. There are many different ways to do this but for our purposes the relatively simple model in agda~\cite{agda} is sufficient (?). Namely the only equality inherent to the type theory is definitional equality, i.e. equality induced by equivalence under $\beta$-reductions and $\eta$-equivalences whereas propositional equality is itself an inductive type (or rather inductive schema) with a single constructor $$\text{refl}_x : x =_A x$$ (the type is parametrized by a type $A$ and $x:A$), i.e. for all types $A$ and $x:A$ we have the rule
\begin{center}
\AxiomC{}
\RightLabel{\scriptsize $=$-intro}
\UnaryInfC{$\text{refl}_x : x =_A x$}
\DisplayProof
\end{center}
Many of the expected properties of equality hold. For instance we can derive the term
\begin{align*}
	\text{subst} &: \Pi_{A}\Pi_{x, y:A}\Pi_{P:A\to Type}\Pi_{e : x=_Ay}\Pi_{p:P x}P y\\
	\text{subst} &\: A\:x\:y\:P\: \text{refl}_x\: px = px 
\end{align*}
by pattern matching, i.e. matching refl the type checker can unify $x$ and $y$.

However there are some issues with this handling of equality. For example it is impossible to define quotient structures: Let's say we prove prime decomposition for PIDs and want to apply it to some ring. Then the ring will be most likely defined as a setoid-like structure with the equality being a more coarse relation on the set of terms and in particular not equal to the equality above. Our proof-as-programs translation will have to take this into account.

@@

\hspace*{-1cm}
\DisplayProof
}
\vspace{3cm}
\small{
	\AxiomC{}
	\RightLabel{\scriptsize Axiom}
	\UnaryInfC{$\vdash \alpha : \Pi_x\Sigma_y f(y)=x$}
	\RightLabel{\scriptsize $\forall E$}
	\UnaryInfC{$\vdash \alpha(x): \Sigma_y f(y)=x$}
	\AxiomC{}
	\RightLabel{\scriptsize Axiom}
	\UnaryInfC{$\vdash \beta : \Pi_x(P(x)\supset Q(f(x), x))$}
	\RightLabel{\scriptsize $\forall E$}
	\UnaryInfC{$\vdash \beta(y) : P(y)\supset Q(f(y), y)$}
	\AxiomC{}
	\RightLabel{\scriptsize Axiom}
	\UnaryInfC{$\vdash\gamma : \Pi_x P(x)$}
	\RightLabel{\scriptsize $\forall E$}
	\UnaryInfC{$\vdash\gamma(y) : P(y)$}
	\RightLabel{\scriptsize $\supset\hspace{-2pt}E$}
	\BinaryInfC{$\vdash \beta(y)(\gamma(y)) : Q(f(y), y)$}
	\AxiomC{}
	\RightLabel{\scriptsize u}
	\UnaryInfC{[$\vdash\delta : f(y) = x$]}
	\RightLabel{\scriptsize subst}
	\BinaryInfC{$\vdash \text{subst}(f(y),x,\Pi_tQ(t, y),\delta,\beta(y)(\gamma(y)) : Q(x, y)$}
	\RightLabel{\scriptsize $\supset\hspace{-2pt}I^u$}
	\UnaryInfC{$\delta : f(y)=x\vdash\text{subst}(f(y),x,\Pi_tQ(t, y),\delta,\beta(y)(\gamma(y)) : Q(x, y)$}
	\RightLabel{\scriptsize$\exists I$}
	\UnaryInfC{$\delta : f(y)=x\vdash \langle y, \text{subst}(f(y),x,\Pi_tQ(t, y),\delta,\beta(y)(\gamma(y))\rangle :\Sigma_y Q(x, y)$}
	\RightLabel{\scriptsize $\exists E$}
	\BinaryInfC{$\vdash \textbf{let }\langle y, \delta\rangle = \alpha(x)\textbf{ in }\langle y,\text{subst}(f(y),x,\Pi_tQ(t, y),\delta,\beta(y)(\gamma(y))\rangle: \Sigma_y Q(x, y)$}
	\RightLabel{\scriptsize $\forall I$}
	\UnaryInfC{$\lambda x.\textbf{let }\langle y, \delta\rangle = \alpha(x)\textbf{ in }\langle y,\text{subst}(f(y),x,\Pi_tQ(t, y),\delta,\beta(y)(\gamma(y))\rangle:\Pi_x \Sigma_y Q(x, y)$}

	\hspace*{-1cm}
	\DisplayProof
}
\hfill
\\\\\hspace{2cm}We can reduce the last term to $\lambda x.\langle \alpha(x)_1, \text{subst}(f(\alpha(x)_1),x,\Pi_tQ(t, \alpha(x)_1),\alpha(x)_2,\beta(\alpha(x)_1)(\gamma(\alpha(x)_1))\rangle$.
